---
title: "README"
author: "Giulio G. Cantone"
date: "5/4/2022"
output: github_document
---

This is a guide for the generating a network from a mixture of multy-layers in R.

# Why I need this in my life?

Because...

## Packages that needed to make the grafting to work

In this guide I will install some useful packages through a package called `pacman`. If you already have these packages installed, you can ignore this section.

### Tidyverse is my basic syntax for data wrangling.

```{r eval=FALSE}
pacman::p_install("tidyverse")
```

### Tidygraph is a syntax for networks manipulation that is inspired by tidyverse. It is built on igraph, and some operations could still be performed on igraph.

```{r eval=FALSE}
pacman::p_install("igraph")
```

```{r eval=FALSE}
pacman::p_install("tidygraph")
```

```{r echo=FALSE}
pacman::p_load(tidyverse,tidygraph)
```

# How to generate deterministic or stochastics graphs with igraph and tidygraph

If you feel confident in this stuff, just jump to the next section.

### Deterministic graph are exact graphs.

A deterministic graph is a graph that has an exact connecting propriety applied over a number of nodes. For example a "complete graph" or "clique" has the propriety that each node of the graph is connected to each other.

In `igraph`, to construct a complete graph, you write this code:

```{r}
igraph::make_full_graph(10, directed = FALSE) -> g1
g1

g1 %>% plot()
```
## Stochastic graphs have a random distributions of edges, following a model.

So, if you randomly generate them, you need to set a "seed" to see always the same outcome. If you don't, there will be a different output each time.

```{r}
set.seed(999)
igraph::random.graph.game(10, p.or.m = .2, directed = FALSE) -> g2
g2

g2 %>% plot()
```

# If you, like me, don't like the igraph format, tidygraph will help you

Deterministic and random graph constructor are easy to recognize in `tidygraph` because deterministic graph generation always starts with the verb `create_`:

```{r}
tidygraph::create_complete(10) -> g1
g1

g1
```
while the random graphs are generated throug the verb `play_`:

```{r}
set.seed(999)
tidygraph::play_erdos_renyi(10, p = .2, directed = FALSE) -> g2
g2
```

You can notice that now the format look like `tibble` from `tidyverse`. This is not a coincidence. This format allows to visualize the attributes of the networks as two relational tables.

Have you noticed that `g1` and `g2` have no columns under the table of the nodes?

One could be mislead to think that this means that those nodes does not exist! But they exists! They are the "invisible" row of the Node Data tibble. If you give a second sight you will see "10 x 0 (active)". This means that there are 10 nodes in the graph.

However, the tibbles of the edges has already 2 attributes: `from` and `to`. These are very particular variables, because usually they cannot be `mutated`. We will learn how to correct manipulate these to engraft our Chimera Network! `from` and `to` also always point to the `rowID` of the nodes, and if `rowID` are re-shuffled, values in `from` and `to` will change to.

## Mutate within tidygraph!

Don't you think that those nodes are a bit hard to recognize?

Tidygraph makes very easy to manipulate attributes in networks:

```{r}
set.seed(999)
g1 %>%
  mutate(
    id =
      cur_group_rows(),
    color = sample(
    c("red", "blue"),
    n(),
    replace = T)
    ) -> g1
g1

g1 %>% plot()
```

Here 2 notes are necessary:

- We gave an `id` to each node as the CURRENT row. This `id` is not related stably to values in `from` and `to`, because these always look for the current `row` of the row, not for a static `id` attribute.
- `plot()` - as ugly as it is - is a very smart guy and it recognize the attribute `color`!

### How can we assign attributes to the links?

Remember the weird wording `(active)`? For each `tidygraph` there is always an active part: the nodes or the edges.

For assigning an attribute to the edges, we need to `activate` the `edges`.

```{r}
set.seed(999)
g1 %>% activate(edges) %>%
  mutate(color = sample(
    c("orange", "aquamarine"),
    n(),
    replace = T,
    p=c(.5,.5)
    )) -> g1
g1

g1 %>% plot()
```

# The essence of Network Grafting:

